\lecture[16]{2024-09-13}{}
A group $G$ is solvable if it has a finite abelian tower, that is,
there exist groups $G_1, \dots, G_m$ such that \[
    G = G_0 \nsup G_1 \nsup \dots G_m = 1,
\] with each $G_i / G_{i+1}$ abelian.

\begin{proposition}
    Let $G$ be a finite group.
    Then an abelian tower admits a cyclic refinement.
\end{proposition}
\begin{proof}
    It suffices to show that any finite abelian group admits a cyclic
    tower (ending in $1$).

    Suppose this holds.
    Let $G$ have abelian tower \[
        G \nsup G_1 \nsup \dots \nsup G_m = 1.
    \] Then $G_{m-1}$ is abelian.

    \TODO[Show that $G_{m-2} \nsup G_{m-1}$ can be refined to a cyclic tower.]
    By induction, the result follows.

    We still need to show that the initial statement holds: any finite
    abelian group admits a cyclic tower.

    The proof is by induction.
    The trivial group is trivial.
    Suppose $\abs G = n > 1$.
    Pick $1 \ne x \in G$.
    Then $G/\angled x$ is finite abelian of order less than $n$.
    If this has a cyclic tower \[
        G/\angled x \nsup H_1 / \angled x
            \nsup \dots \nsup H_m / \angled x = 1,
    \]
    then $G = G_0 \nsup H_1 \nsup \dots \nsup H_m = 1$ is a cyclic
    tower, by the second isomorphism theorem.
\end{proof}

\begin{example}
    Let $G = \GL_n(K)$ where $K$ is a field.
    Let $T$ and $D$ be the subgroups of upper triangular and
    diagonal elements respectively.
    Observe that $D \le T \le G$.
    Let $N$ be the additive subgroup of $M_n(k)$ containing all strictly
    upper triangular elements.

    Define $U = I + N$ and $U_r = I + N^r$ for $r \ge 1$.
    It is easy to check that $U \le G$.
    Recall that the product of two upper triangular matrices is upper
    triangular with $(AB)_{ii} = A_{ii}B_{ii}$.
    Thus the projection map \begin{align*}
        \theta\colon T &\to D \\
        A &\mapsto {(\delta_{ij}A_{ij})}_{ij}
    \end{align*} is a homomorphism.

    Note that $N^r$ consists of all matrices with $0$s in the diagonals
    $j - i \le r$ (induct).
    Thus
    \begin{itemize}
        \item $N^n = \set 0$.
        \item The map \begin{align*}
            \theta_r\colon U_r &\to K^{n-r} \\
            A &\mapsto (a_{1,1+r}, a_{2,2+r}, \dots, a_{n-r,n})
        \end{align*} is a surjective homomorphism,
        with kernel $U_{r+1}$.
    \end{itemize}
\end{example}

$\GL_n(K)$ is not solvable for $n \ge 2$.
$[\GL_n(K), \GL_n(K)] = [\SL_n(K), \SL_n(K)] = \SL_n(K)$.

\subsection{Exact sequences} \label{sec:exact-sequences}
Say we have groups $A_1, A_2, A_3$ with homomorphisms \[
    1 \xrightarrow{\alpha_1} A_1
        \xrightarrow{\alpha_2} A_2
        \xrightarrow{\alpha_3} A_3
        \xrightarrow{\alpha_4} 1.
\] This is an exact sequence of groups iff
$\im \alpha_i = \ker \alpha_{i+1}$ for all $i$.
\begin{align*}
    \alpha_1 &= 1 \mapsto 1 \\
    \alpha_4 &= x \mapsto 1
\end{align*}
Since $1 = \im \alpha_1 = \ker \alpha_2$, we have that $\alpha_2$ is
injective.
Similarly, $\im \alpha_3 = \ker \alpha_4 = A_3$.
Thus $\alpha_3$ is surjective.
We have (upto isomorphism) \begin{align*}
    A_1 &\nsub A_2 \\
    A_2/A_1 &\cong A_3
\end{align*}

\begin{proposition} \label{thm:solve-exact}
    Let $1 \to A \to B \to C \to 1$ be a short exact sequence.
    Then
    \begin{enumerate}
        \item \label{thm:solve-exact:gen}
        $B$ is solvable iff $A$ and $C$ are solvable.
        \item \label{thm:solve-exact:nsub}
        If $H \nsub G$, then $G$ is solvable iff $H$ and
        $G/H$ are solvable.
    \end{enumerate}
\end{proposition}
\begin{proof}
    We already saw that $A \nsub B$ and $C \cong B/A$.
    Thus $\labelcref{thm:solve-exact:nsub} \implies
    \labelcref{thm:solve-exact:gen}$.

    Now since $1 \to H \xrightarrow{i} G \xrightarrow{\pi} G/H \to 1$
    is a short exact sequence, we have that $\labelcref{thm:solve-exact:gen}
    \implies \labelcref{thm:solve-exact:nsub}$.

    We will prove $\labelcref{thm:solve-exact:nsub}$.
    Suppose $G$ is solvable.
    Let $G = G_0 \nsup G_1 \nsup \dots \nsup G_m = 1$ be an abelian
    tower.
    Choose $H_i = H \cap G_i$.
    Then $H = H_0 \nsup H_1 \nsup \dots \nsup H_m = 1$ is an abelian
    tower.
    Also, $G/H = G_0/H \nsup G_1/H \nsup \dots \nsup G_m/H = 1$ is an
    abelian tower.

    Now suppose $H$ and $G/H$ are solvable with towers
    $H = H_0 \nsup H_1 \nsup \dots \nsup H_m = 1$ and
    $G/H = G_0/H \nsup G_1/H \nsup \dots \nsup G_k/H = 1$.
\end{proof}
