\chapter*{The course} \label{chp:course}
\lecture{2024-08-07}{}

\textbf{MS Teams:} 

\section*{Coverage}
\begin{itemize}
    \item Algorithmic paradigms
    \item Flows and applications
    \item \NP-completeness and reductions
    \item Approximation algorithms
    \item Randomized algorithms
    \item Linear programming
\end{itemize}
The last three are dependent on time.

\section*{References}
\begin{itemize}
    \item \textit{Algorithm Design} by Jon Kleinberg and Éva Tardos
    \item \textit{Introduction to Algorithms}
        by Cormen, Leiserson, Rivest, and Stein
    \item \textit{\href{http://jeffe.cs.illinois.edu/teaching/algorithms/}
        {Algorithms}} by Jeff Erickson
\end{itemize}

\section*{Prerequisites}
At least one UG course in algorithms/data structures.
MaTheMaTiCal maTUrItY.

\section*{Evaluation}
\begin{itemize}
    \item[(50\%)] 2 midterms: 3rd/4th week of September and 1st/2nd week of
        November
    \item[(40\%)] Final: end of November
    \item[(10\%)] 7--8 assignments
\end{itemize}

\chapter{Introduction} \label{chp:intro}
\begin{question}
    \textbf{Input:} A list of $n$ numbers $a_1, a_2, \dots, a_n$ and
    an integer $k \le n$.

    \textbf{Output:} The $k$ smallest numbers in the list (in any order).
\end{question}
\begin{itemize}
    \item Sort: $O(n \log n)$
    \item Iterative min-max: Find the minimum, remove it, repeat $k$ times.
        This takes $O(kn)$ time.
    \item Min heap: building a heap takes $O(n)$ time,
        each extract-min takes $O(\log n)$ time,
        so the total time is $O(n + k \log n)$.
    \item Max heap: build a heap from the first $k$ elements in $O(k)$ time.
        For each remaining element, compare it with the maximum element in
        the heap.
        If it is smaller, replace the maximum element with the new element.
        This takes $O(k + (n-k) \log k)$ time.
        Worst case is $O(n \log n)$ when $k = n/2$.
    \item $k$-selection: Find the $k$th smallest element in $O(n)$ time.
        Partition the list around the $k$th element.
        This takes $O(n)$ time.
\end{itemize}

Thus we have the following new question:
\begin{question}
    \textbf{Input:} A list of $n$ numbers $a_1, a_2, \dots, a_n$ and
    an integer $k \le n$.

    \textbf{Output:} The $k$th smallest number in the list.
\end{question}
\begin{solution}
    Select the median of the list in $O(n)$ time.
    Use this as the pivot to partition the list.
    Recurse on the appropriate half.
    \[
        T(n) = n + T(n/2) = O(n) \qedhere
    \]
\end{solution}
This of course requires finding the median in $O(n)$ time.
\begin{question}
    \textbf{Input:} A list of $n$ numbers $a_1, a_2, \dots, a_n$ and an
    integer $k \le n$.

    \textbf{Output:} The median element.
\end{question}
\begin{solution}
    
\end{solution}
