\section{Bootstrapping} \label{sec:boot}
\lecture{2024-09-04}{Bootstrapping}
Start with a basic data structure and keep obtaining an improved data
structure.
Our goal is to obtain, for each $\eps > 0$, a data structure for
2D orthogonal range reporting that consumes $O(n)$ space and processes
queries in $O(n^\eps + k)$ time.

\begin{lemma*} \label{thm:boot}
    Suppose there is a structure $\mcT$ which solves 2D orthogonal range
    reporting with space $S(n)$ and query time $Q(n) + O(k)$.
    Then, for any integer $\lambda \in [2, \frac{n}{2}]$, there exists a
    structure which uses \begin{align*}
        \text{space} &\le \lambda S\ab(\ceil*{\frac n\lambda}) + O(n) \\
        \text{query time} &\le 2 Q\ab(\ceil*{\frac n\lambda})
            + \lambda O\ab(\log\frac n\lambda) + O(k)
    \end{align*}
\end{lemma*}
\begin{proof}
    Divide the points $P$ into $\lambda$ vertical slabs
    $P_1, P_2, \dots, P_\lambda$ of size $\le \ceil{\frac{n}{\lambda}}$.
    Build a structure $\mcT_i$ for each slab $P_i$.
    Furthermore, sort the points in each slab by their $y$-coordinates.

    To query, we locate the overlapping slabs $P_i, P_{i+1}, \dots, P_{i'}$.
    We query $P_i$ and $P_{i'}$ using $\mcT_i$ and $\mcT_{i'}$,
    and for each $P_j$ in between, we query using the sorted list.
    This takes \[
        O(\log \lambda) + 2 Q\ab(\ceil*{\frac n\lambda}) +
            \lambda O\ab(\log\frac n\lambda) + O(k)
    \] time.
\end{proof}

\begin{question*}
    Given a set of $n$ points in the plane and an $\eps > 0$, process
    queries of the form $[x_1, x_2] \times [y_1, y_2]$ using
    $O(n)$ space and $O(n^\eps + k)$ query time.
\end{question*}
\begin{solution}
    Use \cref{thm:boot}.
    We know that Kd-trees solve the problem with $O(n)$ space and
    $O(\sqrt n + k)$ query time.
    Choosing $\lambda = n^{1/3}$ gives the existence of a structure
    $D_1$ with \begin{align*}
        \text{query time} &\le 2 O\ab(\ab(\frac n\lambda)^{\frac12})
            + \lambda O(\log \frac n\lambda) + O(k) \\
        &= O(n^{\frac13} + n^{\frac13} \log n^{\frac13} + k) \\
        &= \wtld{O}(n^{\frac13} + k)
    \end{align*} where $\wtld{O}$ hides logarithmic factors.

    We can now choose $\lambda = n^{1/4}$ to obtain a structure $D_2$
    with \[
        \text{query time} \le \wtld O(n^{\frac14} + k).
    \] Iterating this process $1 + \ceil{\frac 1\eps}$ times gives the
    desired structure.
    If $Q(n) = \wtld O(n^{\frac1\mu})$ for some $\mu \in \N^+$, then
    choosing $\lambda = n^{\frac1{\mu+1}}$ gives a structure with
    $Q(n) = \wtld O(n^{\frac1{\mu+1}})$.
    At each iteration, space occupied stays $O(n)$.
\end{solution}
We can analyze the space in terms of both $n$ and $\eps$.
\begin{lemma}
    The space occupied by $D_t$ is $O(nt)$.
\end{lemma}
\begin{proof}
    The space occupied by $D_1$ is $O(n)$.
    If $S_t(n) \le Cnt$, then \begin{align*}
        S_{t+1}(n) &\le \sum_{i=1}^\lambda S_t(n_i) + C'n \\
        &\le Cnt + C'n \\
        &= Cn\ab(t + \frac{C'}{C}).
    \end{align*} Choosing $C > C'$ gives the claim by induction.
\end{proof}
In our case, we get space $O(n/\eps)$.

\begin{exercise*} \label{thm:boot-ors}
    Generalise this to $d$ dimensions.
\end{exercise*}

\begin{lemma}[Bootstrapping]
    Suppose there is a structure $\mcT$ which solves $d$-dimensional
    orthogonal range reporting with space $S(n)$ and query time
    $Q(n) + O(k)$,
    and a structure $\mcT'$ which solves $(d-1)$-dimensional orthogonal
    range reporting with space $S'(n)$ and query time $Q'(n) + O(k)$.
    Then for each $\lambda \in [2, \frac n2]$,
    there exists an algorithm which solves $d$-dimensional
    range reporting with \begin{align*}
        \text{space} &\le \lambda S\ab(\ceil*{\frac n\lambda})
            + \lambda S'\ab(\ceil*{\frac n\lambda}) \\
        \text{query time} &\le 2 Q\ab(\ceil*{\frac n\lambda})
            + \lambda Q'\ab(\ceil*{\frac n\lambda}) + O(k)
    \end{align*}
\end{lemma}
\begin{proof}
    Divide the points $P$ into $\lambda$ vertical slabs
    $P_1, P_2, \dots, P_\lambda$ of size $\le \ceil{\frac n\lambda}$.
    Build structures $\mcT_i$ and $\mcT'_i$ for each slab $P_i$.

    To query, we locate the overlapping slabs $P_i, P_{i+1}, \dots, P_{i'}$.
    We query $P_i$ and $P_{i'}$ using $\mcT_i$ and $\mcT_{i'}$,
    and for each $P_j$ in between, we query using $\mcT'_j$.
    This takes up \begin{align*}
        \text{space} &\le \lambda (S + S')\ab(\ceil*{\frac n\lambda}) \\
        \text{query time} &\le 2 Q\ab(\ceil*{\frac n\lambda})
            + (\lambda - 2) Q'\ab(\ceil*{\frac n\lambda}) + O(k)
            \qedhere
    \end{align*}
\end{proof}
Note that \cref{thm:boot} is a special case of this for
$d = 2$,
$\mcT'$ being a sorted list,
$S'(n) = O(n)$, and
$Q'(n) = O(\log n)$.

\begin{solution}[Solution to \cref{thm:boot-ors}]
    Suppose that for each $\eps > 0$
    we have a structure $\mcT'$ which solves $(d-1)$-dimensional
    orthogonal range reporting with space $O(n)$ and query time
    $O(n^\eps + k)$.

    We know that Kd-trees solve the problem with $O(n)$ space and
    $O(n^{\frac{d-1}{d}} + k)$ query time, for $d$ dimensions.
    Call this structure $D_0$.

    Fix an $\eps > 0$.
    Using bootstrapping with $\lambda_i = n^{\frac{d-1}{d + (d-1)i}}$,
    we obtain structures $D_1, D_2, \dots$ with \begin{align*}
        S_n(D_i) &= O(n) \\
        Q_n(D_i) &= O(n^{\frac{d-1}{d + (d-1)i} + \frac \eps 2})
    \end{align*}
    The proof is by induction.
    This clearly holds for $D_0$.
    Suppose it holds for $D_{i-1}$, $i \ge 1$.
    Then, \begin{align*}
        S_i(n) &\le \lambda_i S_{i-1}\ab(\ceil*{\frac n{\lambda_i}})
            + \lambda_i O(n) \\
        &= O(n) \\
        Q_i(n) &\le 2 Q_{i-1}\ab(\ceil*{\frac n{\lambda_i}})
            + \lambda_i O(n^{\frac \eps 2}) + O(k) \\
        &= O\ab(\ab(\frac n{\lambda_i})^{\frac{d-1}{d + (d-1)(i-1)} + \frac \eps 2})
            + \lambda_i O(n^{\frac \eps 2}) \\
        &= O\ab(\ab(n^{\frac{d + (d-1)(i-1)}{d + (d-1)i}})
                ^{\frac{d-1}{d + (d-1)(i-1)} + \frac \eps 2})
            + O(n^{\frac{d-1}{d + (d-1)i} + \frac \eps 2}) \\
        &= O\ab(n^{\frac{d-1}{d + (d-1)i} + \frac \eps 2})
    \end{align*}
    For large enough $i$, $Q_i(n) = O(n^\eps)$, giving total query time
    $O(n^\eps + k)$.
    The space remains $O(n)$.

    By induction on $d$, we can obtain such a structure for any
    dimension $d$.
\end{solution}
