\chapter{Point location} \label{chp:point-loc}
\lecture{2024-09-09}{Interval stabbing}

\section{Interval stabbing} \label{sec:stab-1d}
\begin{question*}
    Let $\mcI$ be a set of $n$ intervals in $\R$.
    Process queries of the form $x \in \R$, reporting all intervals which
    contain $x$.
\end{question*}

Look at a special case first.
\begin{question*}
    Let $\mcI$ be a set of $n$ intervals in $\R$ with
    $\bigcap \mcI \ne \O$.
    Process queries of the form $x \in \R$, reporting all intervals which
    contain $x$.
\end{question*}
\begin{solution}
    Let $p \in \bigcap \mcI$, and define \begin{align*}
        \mcI_\ell &= \sset{I \cap (-\infty, p]}{I \in \mcI}, \\
        \mcI_r &= \sset{I \cap [p, +\infty)}{I \in \mcI}.
    \end{align*}
    Note that $\subseteq$ is a total order on each of
    $\mcI_\ell$ and $\mcI_r$.
    Sort both of them in descending order under $\subseteq$.
    WLOG let a query $x$ be to the right of $p$.
    Report from $\mcI_r$ until the first interval
    which does not contain $x$.

    This takes $O(n)$ space and only $O(1 + k)$ query time.
\end{solution}

Generalize this via a binary search tree.
\begin{solution}
    Let $P$ be the set of endpoints of the intervals.
    Let $x_m$ be the median $x$-coordinate in $P$.

    Partition $\mcI$ into $\mcI_\cap$, $\mcI_L$, and $\mcI_R$
    as follows:
    \begin{enumerate}
        \item $\mcI_\cap = \sset{I \in \mcI}{I \ni x_m}$
        \item $\mcI_L = \sset{I \in \mcI}{I \subseteq (-\infty, x_m)}$
        \item $\mcI_R = \sset{I \in \mcI}{I \subseteq (x_m, \infty)}$
    \end{enumerate}
    Use the special case for $\mcI_\cap$
    (this requires storing sorted $\mcI_\cap^\ell$ and $\mcI_\cap^r$).
    Recurse on $\mcI_L$ and $\mcI_R$.

    The height of the tree is $O(\log n)$, since $\mcI_L$ and $\mcI_R$
    both contain at most $n$ endpoints, and hence $n/2$ intervals.
    \[
        H(n) \le 1 + H(n/2) \le \log n + 1
    \] The space requirement is $O(n)$, and the query time is
    $O(\log n + k)$. \begin{align*}
        S(n) &\le O(k_v) + 2S(n/2) \\
        T(n) &\le T(n/2) + O(1 + k_v)
    \end{align*} where $\sum_v k_v = n$.
    The preprocessing time is $O(n \log n)$ (\TODO[exercise]).
\end{solution}

We move on to $2$ dimensions.
\begin{question*}
    Let $\mcR$ be a set of $n$ \textbf{disjoint} rectangles in $\R^2$.
    Process queries of the form $x \in \R$, reporting the rectangle
    containing $x$, if any.
\end{question*}

Again consider the special case.
This time, there is no simple total order.
\begin{question*}
    Let $\mcR$ be a set of $n$ \textbf{disjoint} rectangles in $\R^2$ such
    that $\bigcap \pi_x(\mcR) \ne \O$.
    Process queries of the form $x \in \R$, reporting the rectangle
    containing $x$, if any.
\end{question*}
\begin{solution}
    Let $L$ be a vertical line passing through each rectangle.
    Define \[
        \mcI = \sset{\pi_y(R)}{R \in \mcR}
    \]
    Any two intervals in $\mcI$ are disjoint, since
    $\pi_y(R_1) \cap \pi_y(R_2) \cong (R_1 \cap L) \cap (R_2 \cap L) = \O$
    Thus $\mcI$ can be totally ordered by their lower endpoints.
    Sort it.

    For any query $q = (x, y)$,
    binary search through $\mcI$ to find any interval $I = \pi_y(R) \ni y$.
    At most one such interval exists, and the corresponding rectangle $R$
    is the only possible candidate.

    This takes $O(n)$ space and $O(\log n)$ query time.
\end{solution}

Generalize.
\begin{solution}
    Let $P$ be the set of $x$-endpoints of the rectangles.
    Choose the median.
    Partition $\mcR$ into $\mcR_\cap$, $\mcR_L$, and $\mcR_R$
    as before.
    Use the special case for $\mcR_\cap$.
    Recurse on $\mcR_L$ and $\mcR_R$.

    The height of the tree is $O(\log n)$.
    The space requirement is $O(n)$.
    The query time is $O(\log n)$ at each level, for a total of
    $O(\log^2 n)$.
    However, the $O(\log n)$ is only a binary search, with the same query
    at each node.
    We can apply fractional cascading yet again!
    This gives $O(\log n)$ query time.

    In general, we reduce $d$-dimensional orthogonal point location to
    $d-1$ dimensions.
    The space requirement is $O(n)$ per level, and the query time is
    $O(\log^{d-1} n)$.
\end{solution}
