\lecture{2024-10-14}{}

\begin{examples}
    \item Find the maximum volume of an open rectangular garbage box which
    has surface area $\SI{3}{\meter^3}$.

    Let the dimensions of the box be $x, y, z$,
    with volume $xyz$ and surface area $xy + 2yz + 2zx = 3$.
    Then $z = \frac{3 - xy}{2(x + y)}$, so that the volume \[
        V(x, y) = \frac{xy(3 - xy)}{2(x + y)}.
    \] The domain of interest is $S = \sset{(x, y) \in \R^2}
    {x > 0, y > 0, xy < 3}$.
    This is open and unbounded.
    We'll fix that later.

    For critical points, notice that $V(x, y) = V(y, x)$, and \begin{align*}
        D_1 V(x, y) = 0
            &\implies (3y - 2xy^2)(x + y) - xy(3 - xy) = 0 \\
            &\implies 3y^2 - x^2y^2 - 2xy^3 = 0 \\
            &\implies 3 - x^2 - 2xy = 0.
    \end{align*}
    Symmetrically, $D_2 V(x, y) = 0 \implies 3 - y^2 - 2xy = 0$.
    Thus $x = y$, so $x = y = 1$.
    The volume at $(1, 1)$ is $\frac12$.
    We claim that this is the maximum volume.

    To see this, observe that \[
        V(x,y) = \frac{3xy}{2(x+y)} < \frac32 \min(x, y).
    \] Thus whenever $x < \frac13$ or $y < \frac13$, we have
    $V(x, y) < \frac12$.
    Thus we can restrict our view to the compact set \[
        K = \sset{(x, y) \in \R^2}{x \ge \frac13, y \ge \frac13, xy \le 3}.
    \]
    For $xy = 3$, $z = 0$ so $V(x, y) = 0$.
    For $x = \frac13$ or $y = \frac13$, $V(x, y) < \frac12$.
    Thus the maximum volume is attained at $(x, y) = (1, 1)$ as computed
    above.
\end{examples}

\begin{definition}
    We will denote the set of symmetric $n \times n$ matrices by $S(n)$.
    Alternatively, $S(n)$ is the disjoint union of the sets of
    positive definite, negative definite, indefinite and semi-definite
    matrices.
\end{definition}
\begin{exercise*}
    Let $A \in S(n)$.
    Then
    \begin{enumerate}
        \item $A$ is positive (resp. negative) definite iff all eigenvalues
        of $A$ are strictly positive (resp. negative).
        \item $A$ is indefinite iff $A$ has both positive and negative
        eigenvalues.
        \item $A$ is positive (resp. negative) semi-definite iff all
        eigenvalues of $A$ are non-negative (resp. non-positive)
        and $A$ is singular.
    \end{enumerate}
\end{exercise*}

\begin{exercise*}
    Let $A = \begin{pmatrix}
        a & b \\
        b & c
    \end{pmatrix} \in S(2)$.
    Under what conditions is $A$ positive definite, negative definite,
    indefinite and semi-definite?
\end{exercise*}
